---
title: "Bike Trip Presentation"
author: "Jas Sohi"
date: "Tuesday, January 13, 2015"
output: html_document
---
###Dataset Description: A CSV of bicycle rental transactions from the Bay Area Bike Share Data Challenge.
***
####Problem 1: What was the average total time (in minutes) used by a bicycle in the data?

```{r}
bicycles <- read.csv('bike_trip_data.csv')
names(bicycles)
summary(bicycles$Duration)
avg_duration <- mean(bicycles$Duration)
```

#####The average total time in minutes was `r avg_duration` minutes. 
***
####Problem 2: What was the most popular day by trip frequency in this dataset?
```{r}
bicycles["DayofWeek"] <- as.Date(bicycles$Start.Date, format = "%m/%d/%y %H:%M")
bicycles["DayofWeek"] <- weekdays(bicycles$Day)
table(bicycles["DayofWeek"])
#max(table(bicycles["DayofWeek"]))
popular_dayofweek <- sort((table(bicycles["DayofWeek"])),decreasing = TRUE)[1]
#Most popular day of the week
popweekday <- names(popular_dayofweek)

bicycles["Day"] <- as.Date(bicycles$Start.Date, format = "%m/%d/%y %H:%M")
popular_day <- sort((table(bicycles["Day"])),decreasing = TRUE)[1]
#Most popular date
popdate <- names(popular_day) 
#2013-09-25 was most popular day to start a rental
```

#####If you look at it by day of week, then `r popweekday` was the most popular day of the week. If you look at it by the date, then the most popular day in which bike rentals were started was `r popdate`
***
####Problem 3 (harder): Assuming there are 30 bikes per station, find what date and time the bikes FIRST need to be rebalanced. As in, there are 0 bikes at a terminal for a customer to rent. 

####Problem 3 (easier): Assuming there are 30 bikes per station, find what date the bikes FIRST need to be rebalanced. As in, there are 0 bikes at a terminal for a customer to rent. Do this ignoring "Start.Date" and "End.Date" columns.

```{r}
#needed for aggregate functions later on
bicycles["start_count"] = 1; bicycles["end_count"] = 1

#Aggregate the number of bikes leaving station by day
start_columns <- c("Day","Start.Station","start_count")
dfstart <- bicycles[start_columns]
#rename to station for later merge
names(dfstart)[2] <- "Station"
aggdata <- aggregate(start_count ~.,data = dfstart, sum)

#Aggregate the number of bikes arriving back at the station by day
end_columns <- c("Day","End.Station","end_count")
dfend <- bicycles[end_columns]
#rename to station for later merge
names(dfend)[2] <- "Station"
aggdata2 <- aggregate(end_count ~.,data = dfend, sum)

#Merge the two dataframes
total <- merge(aggdata,aggdata2,by=c("Day","Station"))
total$difference <- total$end_count - total$start_count
#bikes available at start of the day - (add 30)
total$bikes_net30 <- ave(total$difference, total$Station, FUN=function(x) cumsum(c(0, head(x, -1))))
#sort to view by station and date to see what bikes_net30 shows
total <- total[order(total$Station,total$Day),]
final <- total[total$bikes_net30 == -30,] 
#sort ascending by date to find the earliest date that there was a net -30 balance of bikes at any station
final <- final[order(final$Day),]
rebalance_day <- final$Day[1]
#2013-09-05
```

#####I chose the easier option this time (it is my first R project this year :) ). The day that the bikes will first need to be rebalanced is `r rebalance_day`